# Notes and sources {#sec-notes-and-sources}

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("scripts/load-libraries.R"))

source(here::here("scripts/prepare-who-data.R"))

```

## Map basics

### OSM and OSM

[OpenStreetMap](https://www.openstreetmap.org/) data is (c) OpenStreetMap Contributors.

[OpenSeaMap](https://map.openseamap.org) ([description](https://map.openseamap.org/legend.php?lang=en&page=help-website)) uses data from OpenStreetMap and other sources.


### GEBCO gridded bathymetic

The General Bathymetric Chart of the Oceans (GEBCO) <https://www.gebco.net/about_us/overview/>

GEBCO_22_Jun_2024_4b31746b7734.zip downloaded 2024-06-22 from <https://download.gebco.net>


## Global distribution of coral reefs

<https://pacificdata.org/data/dataset/global-distribution-of-coral-reefs> pointed me to <https://data.unep-wcmc.org/datasets/1> where I downloaded the data on 2024-06-24.

LAST PAGE UPDATE: 15/02/23

This dataset shows the global distribution of coral reefs in tropical and subtropical regions. It is the most comprehensive global dataset of warm-water coral reefs to date, acting as a foundation baseline map for future, more detailed, work. This dataset was compiled from a number of sources by UNEP World Conservation Monitoring Centre (UNEP-WCMC) and the WorldFish Centre, in collaboration with WRI (World Resources Institute) and TNC (The Nature Conservancy). Data sources include the Millennium Coral Reef Mapping Project (IMaRS-USF and IRD 2005, IMaRS-USF 2005) and the World Atlas of Coral Reefs (Spalding et al. 2001).

UNEP-WCMC, WorldFish Centre, WRI, TNC (2021). Global distribution of warm-water coral reefs, compiled from multiple sources including the Millennium Coral Reef Mapping Project. Version 4.1. Includes contributions from IMaRS-USF and IRD (2005), IMaRS-USF (2005) and Spalding et al. (2001). Cambridge (UK): UN Environment World Conservation Monitoring Centre. Data DOI: https://doi.org/10.34892/t2wk-5t34

### Citations for the separate entities

IMaRS-USF (Institute for Marine Remote Sensing-University of South Florida) (2005). Millennium Coral Reef Mapping Project. Unvalidated maps. These maps are unendorsed by IRD, but were further interpreted by UNEP World Conservation Monitoring Centre. Cambridge (UK): UNEP World Conservation Monitoring Centre

IMaRS-USF, IRD (Institut de Recherche pour le Developpement) (2005). Millennium Coral Reef Mapping Project. Validated maps. Cambridge (UK): UNEP World Conservation Monitoring Centre

Spalding MD, Ravilious C, Green EP (2001). World Atlas of Coral Reefs. Berkeley (California, USA): The University of California Press. 436 pp.

### USE CONSTRAINTS

UNEP-WCMC General Data License.

For commercial use, please contact business-support@unep-wcmc.org.

----

Metadata link points to file WCMC_008_Global_Distribution_of_Coral_Reefs.pdf

Factsheet link points to <https://biodiversitya-z.org/content/warm-water-coral-reef> also warm-water-coral-reef-factsheet.pdf


## Risk of coral bleaching: NOAA Coral Reef Watch

Data source <https://coralreefwatch.noaa.gov/product/vs/data/fiji.txt> downloaded 2024-06-09.

Plots of a number of island nations: <https://coralreefwatch.noaa.gov/product/vs/timeseries/melanesia.php#fiji>

Description for Daily 5km Regional Virtual Stations, Time Series Data, and Graphs (Updated June 8, 2020) <https://coralreefwatch.noaa.gov/product/vs/description.php#ascii>


## Pocket summary statistics

Data source: <https://pacificdata.org/data/dataset/overseas-visitor-arrivals-df-overseas-visitors> downloaded 2024-06-12.


## Visitor arrivals

Data source: <https://pacificdata.org/data/dataset/overseas-visitor-arrivals-df-overseas-visitors> downloaded 2024-06-16. See also RBF data.


## StatsFiji

<https://www.statsfiji.gov.fj>

<https://www.statsfiji.gov.fj/images/documents/Fiji_MICS_2021/Fiji_MICS_2021_Fact_Sheet_-_Preliminary_Report.pdf>


## Fiji Ocean Policy 2020-2030

Fiji Ministry of Economy. (2021). Republic of Fiji National Ocean Policy 2020-2030. Suva, Fiji: Ministry of Economy. 70 p.

Available from <https://fijiclimatechangeportal.gov.fj/ppss/republic-of-fiji-national-ocean-policy-2020-2030/>

Specifically: <http://fijiclimatechangeportal.gov.fj/wp-content/uploads/2021/01/Fiji_NationalOceansPolicy_2020.pdf>


## Reserve Bank of Fiji

A useful ready-to-read summary can be found at <https://www.rbf.gov.fj/category/chart-book/>.

RBF's National Summary Data Page: <https://www.rbf.gov.fj/statistics/nsdp/>

Data sets used in this report:

* Direction of Trade by Major Trading Partner Countries downloaded 2024-08-06 from 
<https://www.rbf.gov.fj/wp-content/uploads/2024/08/8.4-Direction-of-Trade-by-Major-Trading-Partner.xlsx>

* Consumer price index yearly 1991 - 2023 downloaded 2024-08-06 from
<https://www.rbf.gov.fj/wp-content/uploads/2024/08/Copy-of-7.1-Consumer-Price-Index.xlsx>

* Tourism revenue downloaded 2024-08-06 from 
<https://www.rbf.gov.fj/wp-content/uploads/2024/08/5.5-Hotel-Statistics.xlsx>

The RBF data download web pages include the following:

> Disclaimer: Please refer to the Reserve Bank of Fiji Disclaimer available on our website:  <https://www.rbf.gov.fj/disclaimer/>

<br>

## US historical consumer price index

The Federal Reserve Bank of Minneapolis offers historical CPI data at <https://www.minneapolisfed.org/about-us/monetary-policy/inflation-calculator/consumer-price-index-1800->.


## World Health Organization

The WHO Global Health Observatory is available at <https://www.who.int/data/gho>. The available health indicators and data dimensions are enumerated at <https://ghoapi.azureedge.net/api/Indicator> and <https://ghoapi.azureedge.net/api/Dimension>.


## Sources not used

I looked at the following but for various reasons decided not to include them. I offer links here in case someone wishes to explore them.

### PopGIS3 Fiji

<http://fiji.popgis.spc.int/#bbox=21562,8526695,1053531,882426&c=indicator&i=d2_pop2017.t_pop&t=A07&view=map6>

2017 Census by Division

### All SDG data for Fiji

<https://pacificdata.org/data/dataset/sdg-country-package-for-fiji-dv-sdg-fj>

Dataset description: "This is a subset of Sustainable Development Goals (all) Contains available data on Pacific Sustainable Development Indicators selected by the region and endorsed by Forum Leaders in the...". Source: SDG country package for Fiji

### populuation.un.org

<https://population.un.org/wpp/>

The data portal <https://population.un.org/dataportal/> looks interesting, however it was too slow and/or didn't return data in some cases.

Referenced from <https://database.earth/population/fiji>

<br>

## ISO country codes {#sec-iso-country-codes}

@fig-diabetes uses these three-letter country codes.

```{r}
#| label: tbl-iso-country-codes

diabetes_wide |>
  left_join(ISO_3166_1,
            by = join_by(spatial_dim == Alpha_3)
  ) |>
  select(country_code = spatial_dim, Name,  Official_name, region = parent_location, rank_val) |>
  arrange(country_code) |>
  gt() |>
  tab_header(md("**ISO 3166 three-letter country codes explained**")) |>
  tab_options(table.font.size = 10) |>
  sub_missing()

```

<br>
