# Reefs

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("scripts/load-libraries.R"))

source(here::here("scripts/prepare-osm-data.R"))

theme_set(theme_light() +
            theme(panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  plot.title = element_text(size = rel(2))))

options(scipen = 5)

sf_use_s2(FALSE) # following https://github.com/r-spatial/sf/issues/2366

my_caption <- "Plot: Daniel Moul; data: UNEP-WCMC, WorldFish Centre, WRI, TNC (2021) and Open Street Map"

```

Fiji has extensive reefs: mostly fringing reefs with some barrier reefs and coral mounds, and still fewer atolls, the latter mostly in the Lau group in the east of the country's waters.^[<https://www.taylorfrancis.com/chapters/mono/10.1201/9781003287339-15/atolls-fiji-walter-goldberg-eugene-rankey> ]

```{r}
#| label: get-global-reef-data

fname <- here("data/raw/global-dist-coral/14_001_WCMC008_CoralReefs2021_v4_1/01_Data")

d_raw <- st_read(dsn = fname,
                 layer = "WCMC008_CoralReef2021_Py_v4_1",
                 quiet = TRUE)

d_raw_metadata <- st_read(dsn = fname,
                 layer = "Metadata_CoralReefs",
                 quiet = TRUE)

```

```{r}
#| label: fig-fiji-reefs
#| fig-cap: "Fiji reefs"
#| fig-height: 10
#| fig-width: 10
#| warning: false
#| message: false
#| column: page-right

dta_for_plot <- d_raw |>
  st_crop(my_bbox_fj)

plot_alpha <- 0.3

dta_for_plot |>
  ggplot() +
  geom_sf(data = islands$osm_multipolygons,
          color = "darkgrey", fill = "darkgrey", linewidth = 0.001) +
  geom_sf(data = islands$osm_polygons,
          color = "darkgrey", fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = coastline$osm_polygons,
          color = "darkgrey", fill = "darkgrey",
          size = 0.001) +
  geom_sf(fill = "firebrick", color = "firebrick", alpha = 0.3) +
  scale_x_continuous(expand = expansion(mult = c(0,0))) +
  scale_y_continuous(expand = expansion(mult = c(0,0))) +
  theme(panel.background = element_rect(fill = alpha("lightblue", 0.15))) +
  labs(
    title = "Reefs in most of Fiji",
    subtitle = "Fiji west of 180°",
    x = NULL,
    y = NULL,
    caption = my_caption
  )

```

<br>

I make use of data from *Global distribution of warm-water coral reefs, compiled from multiple sources including the Millennium Coral Reef Mapping Project*, a recently published effort to curate a good data set of warm water coral reefs world-wide:

> This dataset was compiled from a number of sources by UNEP World Conservation Monitoring Centre (UNEP-WCMC) and the WorldFish Centre, in collaboration with WRI (World Resources Institute) and TNC (The Nature Conservancy).^[UNEP-WCMC, WorldFish Centre, WRI, TNC (2021). Global distribution of warm-water coral reefs, compiled from multiple sources including the Millennium Coral Reef Mapping Project. Version 4.1. Includes contributions from IMaRS-USF and IRD (2005), IMaRS-USF (2005) and Spalding et al. (2001). Cambridge (UK): UN Environment World Conservation Monitoring Centre. Data DOI: <https://doi.org/10.34892/t2wk-5t34> and <https://data.unep-wcmc.org/datasets/1> ]

```{r}
#| label: fig-top-reef-info-sources
#| fig-cap: "Top sources of global reef data"
#| message: false
#| warning: false

dta_for_plot <- d_raw |>
  st_drop_geometry() |>
  count(METADATA_I) |>
  mutate(METADATA_I = factor(METADATA_I),
         METADATA_I = fct_lump_min(METADATA_I, min = 200, w = n),
         SOURCE_ID = as.character(METADATA_I)
         ) |>
  filter(METADATA_I != "Other") |>
  left_join(d_raw_metadata |>
              select(SOURCE_ID, DC_TITLE) |>
              mutate(SOURCE_ID = as.character(SOURCE_ID)),
            by = "SOURCE_ID") |>
  mutate(source_desc = str_wrap(glue("{SOURCE_ID}: {DC_TITLE}"), 40),
         source_desc = fct_reorder(source_desc, n))
  
n_sources <- nrow(dta_for_plot)

dta_for_plot |>
  ggplot() +
  geom_col(aes(x = n, y = source_desc),
           fill = "firebrick", alpha = 0.3) +
  geom_text(aes(x = n + 100, y = source_desc, label = n),
            hjust = 0) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme(plot.title.position = "plot" # "#0000CD33")) +
  ) +
  labs(
    title = glue("Top {n_sources} sources of reef information"),
    x = "Count of sources",
    y = NULL,
    caption = my_caption
  )

```

<br>

The "Global distribution of warm-water coral reefs" dataset draws most of it's data from the Millennium Coral Reef Mapping Project. The third-most common source was unique to Fiji: *Exposed Reefs* - Fiji Lands Information System, Department of Lands and Survey (Ministry of Lands, Mineral Resources & Environment for Fiji).^[I was not able to find *Exposed Reefs* online.]

```{r}
#| label: tbl-top-reef-data

dta_for_plot |>
  arrange(desc(n)) |>
  select(SOURCE_ID, FREQ = n) |>
  left_join(d_raw_metadata |>
              select(SOURCE_ID, DC_TITLE, DESCRIPTIO, CITATION, PUB_DATE) |>
              mutate(SOURCE_ID = as.character(SOURCE_ID)),
            by = "SOURCE_ID") |>
  rename(DESCRIPTION = DESCRIPTIO) |>
  gt() |>
  tab_header(md("**Top 6 sources**")) |>
  tab_options(table.font.size = 10)

```

<br>

Zooming into the area between the Yasawas and northwestern Viti Levu, one can see fringing reefs, barrier reefs, and coral mounds.

```{r}
#| label: fig-fiji-reefs-yasawas
#| fig-cap: "Fiji reefs in the Yasawas and Mamanucas"
#| fig-height: 10
#| fig-width: 10
#| warning: false
#| message: false
#| column: page-right

dta_for_plot <- d_raw |>
  st_crop(my_bbox_yasawas)

dta_for_plot |>
  ggplot() +
  geom_sf(data = coastline_yasawas$osm_points,
          color = "darkgrey",
          size = 0.001) +
  geom_sf(data = coastline_yasawas$osm_polygons,
          color = "darkgrey", fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = islands_yasawas$osm_polygons,
          color = "darkgrey", fill = "darkgrey",
          size = 0.001) +
  geom_sf(fill = "firebrick", color = "firebrick", alpha = 0.3) +
  scale_x_continuous(expand = expansion(mult = c(0,0))) +
  scale_y_continuous(expand = expansion(mult = c(0,0))) +
  theme(panel.background = element_rect(fill = alpha("lightblue", 0.15))) + # "#0000CD33")) +
  labs(
    title = "Fiji reefs in the Yasawas and Mamanucas",
    x = NULL,
    y = NULL,
    caption = my_caption
  )

```

<br>

Many reefs are named and considered traditionally associated with specific land-owning family groups (mataqali). In Fiji all land below high tide and all ocean area are considered publicly accessible.

```{r}
#| label: fig-fiji-reefs-yasawas-with-names
#| fig-cap: "Fiji reefs in the Yasawas and Mamanucas with reef names"
#| fig-height: 10
#| fig-width: 10
#| warning: false
#| message: false
#| column: page-right

dta_for_plot |>
  ggplot() +
  geom_sf(data = coastline_yasawas$osm_points,
          color = "darkgrey",
          size = 0.001) +
  geom_sf(data = coastline_yasawas$osm_polygons,
          color = "darkgrey", 
          fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = islands_yasawas$osm_polygons,
          color = "darkgrey", 
          fill = "darkgrey",
          size = 0.001) +
  geom_sf(fill = "firebrick", color = "firebrick", alpha = 0.3) +
  geom_sf_text(aes(label = if_else(NAME != "Not Reported",
                                   NAME,
                                   NA)), 
                   size = 2, color = "black",
               check_overlap = TRUE,
               na.rm = TRUE) +
  scale_x_continuous(expand = expansion(mult = c(0,0))) +
  scale_y_continuous(expand = expansion(mult = c(0,0))) +
  theme(panel.background = element_rect(fill = alpha("lightblue", 0.15))) + # "#0000CD33")) +
  labs(
    title = "Fiji reefs in the Yasawas and Mamanucas",
    subtitle = "With reef names (overlaping names omitted)",
    x = NULL,
    y = NULL,
    caption = my_caption
  )

```

<br>

Zooming in still further, Nacula is the island in the Yasawas where the resort I'm visiting is located.

```{r}
#| label: fig-fiji-reefs-yasawas-nacula
#| fig-cap: "Fiji reefs around Nacula in the Yasawas. Reef data from Global Reef Distribution. Other data from Open Streat Map."
#| fig-height: 10
#| fig-width: 10
#| warning: false
#| message: false
#| column: page-right
  
dta_for_plot  |>
  st_crop(my_bbox_nacula) |>
  ggplot() +
  geom_sf(data = coastline_yasawas$osm_points |>
            st_crop(my_bbox_nacula),
          color = "darkgrey",
          size = 0.001) +
  geom_sf(data = coastline_yasawas$osm_polygons |>
            st_crop(my_bbox_nacula),
          color = "darkgrey", 
          fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = islands_yasawas$osm_polygons |>
            st_crop(my_bbox_nacula),
          color = "darkgrey", 
          fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = nacula_places$osm_polygons |>
            st_crop(my_bbox_nacula),
          color = "blue", shape = 15,
          size = 2) +
  geom_sf_text(data = nacula_places$osm_points |>
                 st_crop(my_bbox_nacula),
               aes(label = name),
               color = "black", hjust = 0,
               size = 3) +
  geom_sf(fill = "firebrick", color = "firebrick", alpha = 0.3) +
  scale_x_continuous(expand = expansion(mult = c(0,0))) +
  scale_y_continuous(expand = expansion(mult = c(0,0))) +
  theme(panel.background = element_rect(fill = alpha("lightblue", 0.15))) + # "#0000CD33")) +
  labs(
    title = "Fiji reefs around Nacula island in the Yasawas",
    subtitle = "Reef data from Global Reef Distribution, other from OSM",
    x = NULL,
    y = NULL,
    caption = my_caption
  )

```

<br>

As one might expect, a project with the goal of collecting Global Reef Distribution data provides higher resolution and likely more accurate data (@fig-fiji-reefs-yasawas-nacula above) compared to what's available from the Open Street Map (@fig-fiji-reefs-yasawas-nacula-osm below). Nonetheless, it's impressive to me that OSM has this much data about Fiji.

```{r}
#| label: fig-fiji-reefs-yasawas-nacula-osm
#| fig-cap: "Fiji reefs around Nacula in the Yasawas. All data from Open Streat Map."
#| fig-height: 10
#| fig-width: 10
#| warning: false
#| message: false
#| column: page-right

ggplot() +
  geom_sf(data = coastline_yasawas$osm_points |>
            st_crop(my_bbox_nacula),
          color = "darkgrey",
          size = 0.001) +
  geom_sf(data = coastline_yasawas$osm_polygons |>
            st_crop(my_bbox_nacula),
          color = "darkgrey", 
          fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = islands_yasawas$osm_polygons |>
            st_crop(my_bbox_nacula),
          color = "darkgrey", 
          fill = "darkgrey",
          size = 0.001) +
  geom_sf(data = nacula_reefs$osm_multipolygons |>
            st_shift_longitude() |>
            st_crop(my_bbox_nacula),
          color = "firebrick", fill = "firebrick", alpha = 0.3) +
  geom_sf(data = nacula_reefs$osm_points |>
            st_shift_longitude() |>
            st_crop(my_bbox_nacula),
          color = "firebrick", size = 0.01, alpha = 0.3) +
  geom_sf(data = nacula_places$osm_polygons |>
            st_crop(my_bbox_nacula),
          color = "blue", shape = 15,
          size = 2) +
  geom_sf_text(data = nacula_places$osm_points |>
                 st_crop(my_bbox_nacula),
               aes(label = name),
               color = "black", hjust = 0,
               size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0,0))) +
  scale_y_continuous(expand = expansion(mult = c(0,0))) +
  theme(panel.background = element_rect(fill = alpha("lightblue", 0.15))) + # "#0000CD33")) +
  labs(
    title = "Fiji reefs around Nacula island in the Yasawas",
    subtitle = "Data from Open Street Map",
    x = NULL,
    y = NULL,
    caption = my_caption
  )

```

<br>
