# WHO indicators

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("scripts/load-libraries.R"))

source(here::here("scripts/prepare-who-data.R"))

theme_set(theme_light() +
            theme(panel.background = element_blank(),
                  panel.grid = element_blank(),
                  panel.border = element_blank()))

```

The World Health Organization tracks many health indicators at the country and region level and makes the data available at the [Global Health Observatory](https://www.who.int/data/gho). Below I look at healthy life expectancy and some of the most significant non-communicable diseases (NCDs).


## Healthy life expectancy at birth (HALE)

Healthy life expectancy at birth is the "average number of years that a person can expect to live in “full health” from birth."^[<https://data.who.int/indicators/i/48D9B0C/C64284D> ].

There are some common patterns

1. Women have a longer HALE than men in most countries.

2. The COVID-19 pandemic reduced HALE for most countries and both sexes.

3. "Worldwide, healthy life expectancy at birth (years) has improved by 3.57 years from 58.3 years in 2000 to 61.9 years in 2021."^[ibid.]. 

Unfortunately Fiji did not experience the same level of improvements. These patterns are visible in @fig-hale-3-countries, which compares Fiji, France and USA.

```{r}
#| label: fig-hale-3-countries
#| fig-cap: "Comparing Fiji's Healthy life expectancy at birth (HALE) to France and USA"
#| fig-height: 6
#| fig-width: 8

hale |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = dim1),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Healthy life expectancy at birth: Fiji, France, USA",
    x = NULL,
    y = "Age",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

Fiji is in the upper middle income group as defined by the World Bank (WB_UMI: GNI per capita USD 4,516 TO 14,005 in 2023).^[World Bank Country and Lending Groups <https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups> ] In 2000, Fiji was already a lagging HALE outlier, and while the upper middle income group improved over two decades, Fiji's HALE remained essentially unchanged, making the gap even larger (@fig-hale-fiji-vs-wb-income-group).

```{r}
#| label: fig-hale-fiji-vs-wb-income-group
#| fig-cap: "HALE: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- hale |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 2000],
         low_norm = low / low[time_dim == 2000],
         high_norm = high / high[time_dim == 2000],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = spatial_dim),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  labs(
    subtitle = "A: Fiji is an outlier in the UMI group",
    x = NULL,
    y = "Age",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group",
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low_norm, ymax = high_norm, fill = spatial_dim),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    subtitle = "B: Relative to year 2000 = 100%",
    x = NULL,
    y = "Relative improvement in age",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Fiji has not been improving in healthy life expectancy at birth (HALE)",
                 "\nat a rate similar to any of the World Bank income groups"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

# TODO: Possibly plot other WB_UMI countries together with Fiji

# TODO: Possibly plot other PICTs together with Fiji

```

<br>

## Non-communicable disease (NDC) mortality rate

WHS2_131 Age-standardized NCD mortality rate (per 100 000 population)

TODO: remove text above

```{r}
#| label: prepare-ncd-mortality-data

ncd_long <- ncd_mortality |>
  pivot_longer(cols = dim1,
               values_to = "sex_value")

ncd_wide <- ncd_long |>
  pivot_wider(names_from = sex_value,
              values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

```

<br>

At the regional level (@fig-ncd-region), the western Pacific suffers the worst NCD mortality rates (panel A), and Fiji has one of the highest rates (panel B).

```{r}
#| label: fig-ncd-region
#| fig-cap: "NCD mortaility 2019 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- ncd_long |>
  filter(spatial_dim_type == "COUNTRY",
         sex_value == "SEX_BTSX",
         time_dim == 2019)

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,    ~y,
  "Africa",                  720,   0.0039,
  "Americas",                400,   0.0042,
  "Eastern Mediterranean",   800,   0.002,
  "Europe",                  300,   0.003,
  "South-East Asia",         750,   0.003,
  "Western Pacific",         900,   0.0012,
)

ecdf_fun <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Western Pacific") |>
    pull(numeric_value)
)

fiji_value <- dta_for_plot |>
    filter(parent_location == "Western Pacific",
           spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(parent_location == "Americas",
           spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)
  
ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- dta_for_plot |>
  ggplot() +
  geom_density(aes(x = numeric_value, color = parent_location),
               linewidth = 0.6, alpha = 0.8,
               show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4,
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(#labels = label_percent(),
    expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "People per 100K",
    y = NULL
  )

p2 <- dta_for_plot |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 40, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "People per 100K",
    y = "Percent of all countries in region"
    
  )

p1 + p2 +
  plot_annotation(
    title = glue("Age-standardized NCD mortality, per 100,000 people (2019)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

```

<br>

```{r}
#| label: fig-ncd-fiji-vs-wb-income-group
#| fig-cap: "Non-communicable disease mortality rates: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- ncd_mortality |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 2000],
         low_norm = low / low[time_dim == 2000],
         high_norm = high / high[time_dim == 2000],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim),
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "A: Rate",
    x = NULL,
    y = "Rate",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "B: Year 2000 = 100%",
    x = NULL,
    y = "Relative improvement",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("During this period Fiji started out lowering NDC mortality at a rate similar to",
                 "\nWorld Bank high income group, then around 2004 the rate of improvement slowed markedly (panel B)."),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

```{r}
#| label: fig-ncd-mortality-3-countries
#| fig-cap: "Comparing Fiji's Healthy life expectancy at birth (HALE) to France and USA"
#| fig-height: 6
#| fig-width: 8

ncd_mortality |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.002, 0))) +
  expand_limits(y = 0) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "NCS mortality rates per 100,000 people: Fiji, France, USA",
    x = NULL,
    # y = "Age",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

## Diabetes

```{r}
#| label: prepare-diabetes-data

# Age-standardized death rates, diabetes mellitus, per 100,000

diabetes_long <- diabetes |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

diabetes_wide <- diabetes_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

# TODO: why only 2004 data?

```

<br>

The following uses 2004 data, which was the latest year available in the GHO data (at least, it's the latest I could find).

At the regional level, Africa is the only region in which the bulk of the countries are suffering a higher rate, while the Americas have the widest range and the worst rates world-wide (@fig-diabetes-region).

```{r}
#| label: fig-diabetes-region
#| fig-cap: "Diabetes 2004 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,    ~y,
  "Africa",                  65,   0.027,
  "Americas",                90,   0.007,
  "Eastern Mediterranean",   30,   0.023,
  "Europe",                  15,   0.065,
  "South-East Asia",         36,   0.05,
  "Western Pacific",         58,   0.015,
)

dta_for_plot <- diabetes_long |>
  filter(sex_value == "SEX_BTSX")

ecdf_fun <- ecdf(
  dta_for_plot |>
  filter(parent_location == "Western Pacific") |>
  pull(numeric_value)
  )

fiji_value <- dta_for_plot |>
    filter(spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)

ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- diabetes_long |>
  ggplot() +
   geom_density(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4, 
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "People per 100K",
    y = NULL
  )

p2 <- diabetes_long |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 5, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "People per 100K",
    y = "Percent of all countries in region"
    
  )

p1 + p2 +
  plot_annotation(
    title = glue("Age-standardized death rates, diabetes mellitus, per 100,000 people (2004)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

# TODO: Why isn't the USA point actually on the Americas line?

```

<br>

Fiji ranked 112 out of 191 countries (59th percentile) in death from diabetes (@fig-diabetes). 

```{r}
#| label: fig-diabetes
#| fig-cap: "Diabetes 2004"
#| fig-height: 16
#| fig-width: 10

rank_min <- min(diabetes_wide$rank_val)
rank_max <- max(diabetes_wide$rank_val)

diabetes_wide |>
  ggplot() +
  geom_errorbarh(aes(y = spatial_dim_label, xmin = SEX_MLE, , xmax = SEX_FMLE, color = parent_location),
            linewidth = 0.1, height = 0, alpha = 0.8,
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_BTSX, color = parent_location),
            size = 1, alpha = 0.8,
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_MLE, color = parent_location),
            size = 3, alpha = 0.8, shape = "M",
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_FMLE, color = parent_location),
            size = 3, alpha = 0.8, shape = "F",
            show.legend = FALSE) +
  # scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  facet_wrap(~parent_location, scales = "free_y") +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    title = glue("Age-standardized death rates, diabetes mellitus, per 100,000 people (2004)"),
    subtitle = glue("With country rank lowest (1) to highest ({rank_max}) death rate",
                    "\nShowing M, F, and both sexes (the later with a point)"),
    x = NULL,
    y = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

The following three-letter country codes are used in @fig-diabetes above.

```{r}
diabetes_wide |>
  left_join(ISO_3166_1,
            by = join_by(spatial_dim == Alpha_3)
  ) |>
  select(country_code = spatial_dim, Name,  Official_name, region = parent_location, rank_val) |>
  arrange(country_code) |>
  gt() |>
  tab_header(md("**ISO 3166 three-letter country codes explained**")) |>
  tab_options(table.font.size = 10) |>
  sub_missing()

```

<br>

## Prevalence of obesity among adults

NCD_BMI_30A Prevalence of obesity among adults, BMI &GreaterEqual; 30 (age-standardized estimate) (%)

```{r}
#| label: prepare-obesity-data

obesity_long <- bmi_obesity |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

obesity_wide <- obesity_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

```

<br>

At the regional level ...

```{r}
#| label: fig-obesity-region
#| fig-cap: "Adult obesity 2022 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- obesity_long |>
  filter(spatial_dim_type == "COUNTRY",
         sex_value == "SEX_BTSX",
         time_dim == 2022)

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,     ~y,
  "Africa",                  15,    0.05,
  "Americas",                37,    0.042,
  "Eastern Mediterranean",   44,    0.028,
  "Europe",                  20.5,  0.072,
  "South-East Asia",         12,    0.078,
  "Western Pacific",         53,    0.014,
)

ecdf_fun <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Western Pacific") |>
    pull(numeric_value)
)

fiji_value <- dta_for_plot |>
    filter(parent_location == "Western Pacific",
           spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(parent_location == "Americas",
           spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)

ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- dta_for_plot |>
  ggplot() +
  geom_density(aes(x = numeric_value, color = parent_location),
               linewidth = 0.6, alpha = 0.8,
               show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4,
            show.legend = FALSE) +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "Density",
    x = "Percent of adult population",
    y = NULL
  )

p2 <- dta_for_plot |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 3, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "Cumulative Distribution",
    x = "Percent of adult population",
    y = "Percent of all countries in region"
    
  )

p1 + p2 +
  plot_annotation(
    title = glue("Adult obesity (2019)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

```

<br>

```{r}
#| label: fig-obesity-fiji-vs-wb-income-group
#| fig-cap: "Adult obesity: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- bmi_obesity |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 1990],
         low_norm = low / low[time_dim == 1990],
         high_norm = high / high[time_dim == 1990],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  # geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = spatial_dim),
  #             alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim),
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(#labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  expand_limits(y = 0) +
  labs(
    x = NULL,
    y = "Rate",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p2 <- dta_for_plot |>
  ggplot() +
  # geom_ribbon(aes(x = time_dim, ymin = low_norm, ymax = high_norm, fill = spatial_dim),
  #             alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "Year 1990 = 100%",
    x = NULL,
    y = "Relative change",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Adult obesity in Fiji has grown more slowly than any World Bank income groups",
                 "\nalbeit from a higher starting percentage"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

```{r}
#| label: fig-obesity-3-countries
#| fig-cap: "Comparing Fiji's adult obesity to France and USA"
#| fig-height: 6
#| fig-width: 8

bmi_obesity |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0.002, 0.02))) +
  expand_limits(y = 0) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Fiji, France, USA: Adult obseity",
    x = NULL,
    y = "Adult obseity",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

## Mean BMI (kg/m²)

NCD_BMI_MEAN Mean BMI (kg/m²) (age-standardized estimate)

```{r}
#| label: prepare-mean-bmi-data

# NCD_BMI_MEAN Mean BMI (kg/m²) (age-standardized estimate)

bmi_mean_long <- bmi_mean |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

bmi_mean_wide <- bmi_mean_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

```

<br>

At the regional level ...

```{r}
#| label: fig-bmi-mean-region
#| fig-cap: "Mean BMI 2016 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- obesity_long |>
  filter(spatial_dim_type == "COUNTRY",
         sex_value == "SEX_BTSX",
         time_dim == 2016)

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,     ~y,
  "Africa",                  12,    0.065,
  "Americas",                33,    0.055,
  "Eastern Mediterranean",   42,    0.028,
  "Europe",                  18.5,  0.085,
  "South-East Asia",         10,    0.1,
  "Western Pacific",         50,    0.016,
)

ecdf_fun <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Western Pacific") |>
    pull(numeric_value)
)

fiji_value <- dta_for_plot |>
    filter(parent_location == "Western Pacific",
           spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(parent_location == "Americas",
           spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)

ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- dta_for_plot |>
  ggplot() +
  geom_density(aes(x = numeric_value, color = parent_location),
               linewidth = 0.6, alpha = 0.8,
               show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4,
            show.legend = FALSE) +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "Percent of population",
    y = NULL
  )

p2 <- dta_for_plot |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 3, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  expand_limits(y = 0) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "Percent of population",
    y = "Percent of all countries in region"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Mean BMI (2016)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

```

TODO: Does @fig-bmi-mean-region make sense compared to adult obesity (@fig-obesity-region)

<br>

```{r}
#| label: fig-bmi-mean-fiji-vs-wb-income-group
#| fig-cap: "Mean MBI: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- bmi_mean |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 1975],
         low_norm = low / low[time_dim == 1975],
         high_norm = high / high[time_dim == 1975],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim),
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  expand_limits(y = 0) +
  labs(
    x = NULL,
    y = "Rate",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "Year 1975 = 100%",
    x = NULL,
    y = "Relative improvement",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Mean BMI in Fiji has grown even faster",
                 " than World Bank lower and lower middle income groups"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

```{r}
#| label: fig-bmi-mean-3-countries
#| fig-cap: "Comparing Fiji's mean BMI to France and USA"
#| fig-height: 6
#| fig-width: 8

bmi_obesity |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  # geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = dim1),
  #             alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(#labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0.002, 0.02))) +
  expand_limits(y = 0) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Fiji, France, USA: Age adjusted mean BMI",
    x = NULL,
    y = "Mean BMI",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

TODO: Does @fig-bmi-mean-3-countries make sense compared to adult obesity (@fig-obesity-region)

<br>
