# WHO indicators

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("scripts/load-libraries.R"))

source(here::here("scripts/prepare-who-data.R"))

theme_set(theme_light() +
            theme(panel.background = element_blank(),
                  panel.grid = element_blank(),
                  panel.border = element_blank()))

```

## Healthy life expectancy at birth (HALE)

```{r}
#| label: fig-hale-3-countries
#| fig-cap: "Comparing Fiji's Healthy life expectancy at birth (HALE) to France and USA"
#| fig-height: 6
#| fig-width: 8

hale |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = dim1),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Fiji vs USA: Healthy life expectancy at birth (HALE)",
    x = NULL,
    y = "Age",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

```{r}
#| label: fig-hale-fiji-vs-wb-income-group
#| fig-cap: "HALE: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 8

dta_for_plot <- hale |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 2000],
         low_norm = low / low[time_dim == 2000],
         high_norm = high / high[time_dim == 2000],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = spatial_dim),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  # facet_wrap(~spatial_dim) +
  labs(
    # title = glue("Fiji has not been improving in 'healthy life expectancy at birth' (HALE)",
    #              "\ncompared to World Bank income groups"),
    x = NULL,
    y = "Age",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group",
    # caption = "Daniel Moul. Source: WHO GHO"
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low_norm, ymax = high_norm, fill = spatial_dim),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent()) +
  # facet_wrap(~spatial_dim) +
  labs(
    subtitle = "Year 2000 = 100%",
    x = NULL,
    y = "Relative improvement in age",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Fiji has not been improving in 'healthy life expectancy at birth' (HALE)",
                 "\ncompared to World Bank income groups"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

TODO: confirm which WB income group Fiji is in

TODO: Plot other PICTs together with Fiji

## Diabetes

```{r}

# Age-standardized death rates, diabetes mellitus, per 100,000

diabetes_long <- diabetes |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

diabetes_wide <- diabetes_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

# why only 2004 data?

```

```{r}
#| label: fig-diabetest
#| fig-cap: "Diabetes 2004"
#| fig-height: 16
#| fig-width: 10

rank_min <- min(diabetes_wide$rank_val)
rank_max <- max(diabetes_wide$rank_val)

diabetes_wide |>
  ggplot() +
  geom_errorbarh(aes(y = spatial_dim_label, xmin = SEX_MLE, , xmax = SEX_FMLE, color = parent_location),
            linewidth = 0.1, height = 0, alpha = 0.8,
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_BTSX, color = parent_location),
            size = 1, alpha = 0.8,
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_MLE, color = parent_location),
            size = 3, alpha = 0.8, shape = "M",
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_FMLE, color = parent_location),
            size = 3, alpha = 0.8, shape = "F",
            show.legend = FALSE) +
  # scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  facet_wrap(~parent_location, scales = "free_y") +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    title = glue("Age-standardized death rates, diabetes mellitus, per 100,000 people"),
    subtitle = glue("With country rank lowest (1) to highest ({rank_max}) death rate",
                    "\nShowing M, F, and both sexes (the later with a point)"),
    x = NULL,
    y = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

```{r}
diabetes_wide |>
  left_join(ISO_3166_1,
            by = join_by(spatial_dim == Alpha_3)
  ) |>
  select(country_code = spatial_dim, parent_location, Name, Official_name) |>
  arrange(country_code) |>
  mutate(idx = row_number()) |>
  gt() |>
  tab_header(md("**ISO 3166 three-letter country codes explained**")) |>
  tab_options(table.font.size = 10) |>
  sub_missing()

```

<br>

## Other indicators of interest

Data available

* NCD_BMI_MEAN Mean BMI (kg/m²) (age-standardized estimate)
* NCD_BMI_30A Prevalence of obesity among adults, BMI &GreaterEqual; 30 (age-standardized estimate) (%)
* NCD_HYP_PREVALENCE_A Prevalence of hypertension among adults aged 30-79 years, age-standardized
* WHS2_131 Age-standardized NCD mortality rate  (per 100 000 population)

Data not available

* WHS2_161 Age-standardized mortality rate by cause (per 100 000 population) - Cardiovascular
* WHS2_160 Age-standardized mortality rate by cause (per 100 000 population) - Cancer

<br>

TODO: Add more visualizations here. Should any plots be functions instead?

```{r}

```

