# WHO indicators

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("scripts/load-libraries.R"))

source(here::here("scripts/prepare-who-data.R"))

theme_set(theme_light() +
            theme(panel.background = element_blank(),
                  panel.grid = element_blank(),
                  panel.border = element_blank()))

```

The World Health Organization tracks many health indicators at the country and region level and makes the data available at the [Global Health Observatory](https://www.who.int/data/gho). Below I look at healthy life expectancy and some aspects of non-communicable diseases (NCDs), which disproportionally affect the people of Fiji.


## Healthy life expectancy at birth (HALE)

Healthy life expectancy at birth is the "average number of years that a person can expect to live in 'full health' from birth."^[<https://data.who.int/indicators/i/48D9B0C/C64284D> ].

There are some common patterns across countries

1. Women have a longer HALE than men in most countries.

2. The COVID-19 pandemic reduced HALE for most countries and both sexes.

3. "Worldwide, healthy life expectancy at birth (years) has improved by 3.57 years from 58.3 years in 2000 to 61.9 years in 2021."^[ibid.]. 

Unfortunately Fiji did not experience the same level of improvements. These patterns are visible in @fig-hale-3-countries, which compares Fiji, France and USA.

Fiji is in the upper middle income group as defined by the World Bank (WB_UMI: GNI per capita USD 4,516 TO 14,005 in 2023).^[World Bank Country and Lending Groups <https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups> ] In 2000, Fiji was already a lagging HALE outlier, and while the upper middle income group improved over two decades, Fiji's HALE remained essentially unchanged, making the gap even larger (@fig-hale-fiji-vs-wb-income-group).

```{r}
#| label: fig-hale-fiji-vs-wb-income-group
#| fig-cap: "HALE: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- hale |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 2000],
         low_norm = low / low[time_dim == 2000],
         high_norm = high / high[time_dim == 2000],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = spatial_dim),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  labs(
    subtitle = "A: Fiji is an outlier in the UMI group",
    x = NULL,
    y = "Age",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group",
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low_norm, ymax = high_norm, fill = spatial_dim),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    subtitle = "B: Relative to year 2000 = 100%",
    x = NULL,
    y = "Relative improvement in age",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Fiji has not been improving in healthy life expectancy at birth (HALE)",
                 "\nat a rate similar to any of the World Bank income groups"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

# TODO: Possibly plot other WB_UMI countries together with Fiji

# TODO: Possibly plot other PICTs together with Fiji

```

<br>

On average, women in France have about ten more years of healthy life compared to women in Fiji (@fig-hale-3-countries).

```{r}
#| label: fig-hale-3-countries
#| fig-cap: "Comparing Fiji's Healthy life expectancy at birth (HALE) to France and USA"
#| fig-height: 6
#| fig-width: 8

hale |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_ribbon(aes(x = time_dim, ymin = low, ymax = high, fill = dim1),
              alpha = 0.3) +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Healthy life expectancy at birth: Fiji, France, USA",
    x = NULL,
    y = "Age",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

## Non-communicable disease (NDC) mortality rate

Non-communicable diseases (NDC) are chronic diseases, and many have behavioral factors. The WHO publishes a useful overview [here](https://www.who.int/news-room/fact-sheets/detail/noncommunicable-diseases).

```{r}
#| label: prepare-ncd-mortality-data

# WHS2_131 Age-standardized NCD mortality rate (per 100 000 population)

ncd_long <- ncd_mortality |>
  pivot_longer(cols = dim1,
               values_to = "sex_value")

ncd_wide <- ncd_long |>
  pivot_wider(names_from = sex_value,
              values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

```

<br>

At the regional level (@fig-ncd-region), the western Pacific suffers the worst NCD mortality rates (panel A), and Fiji has one of the highest rates (panel B).

```{r}
#| label: fig-ncd-region
#| fig-cap: "NCD mortaility 2019 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- ncd_long |>
  filter(spatial_dim_type == "COUNTRY",
         sex_value == "SEX_BTSX",
         time_dim == 2019)

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,    ~y,
  "Africa",                  720,   0.0039,
  "Americas",                400,   0.0042,
  "Eastern Mediterranean",   800,   0.002,
  "Europe",                  300,   0.003,
  "South-East Asia",         750,   0.003,
  "Western Pacific",         900,   0.0012,
)

ecdf_fun <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Western Pacific") |>
    pull(numeric_value)
)

fiji_value <- dta_for_plot |>
    filter(parent_location == "Western Pacific",
           spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(parent_location == "Americas",
           spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)
  
ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- dta_for_plot |>
  ggplot() +
  geom_density(aes(x = numeric_value, color = parent_location),
               linewidth = 0.6, alpha = 0.8,
               show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4,
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(#labels = label_percent(),
    expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "People per 100K",
    y = NULL
  )

p2 <- dta_for_plot |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 40, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "People per 100K",
    y = "Percent of all countries in region"
    
  )

p1 + p2 +
  plot_annotation(
    title = glue("Age-standardized NCD mortality, per 100,000 people (2019)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

```

<br>

```{r}
#| label: fig-ncd-fiji-vs-wb-income-group
#| fig-cap: "Non-communicable disease mortality rates: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- ncd_mortality |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 2000],
         low_norm = low / low[time_dim == 2000],
         high_norm = high / high[time_dim == 2000],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim),
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "A: Rate",
    x = NULL,
    y = "Rate",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "B: Year 2000 = 100%",
    x = NULL,
    y = "Relative improvement",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("During this period Fiji started out lowering NDC mortality at a rate similar to World Bank high income group (panel A)",
                 "\nthen around 2004 the rate of improvement slowed markedly (panel B)."),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

The NCD rate for males in Fiji has been the most stubborn (@fig-ncd-mortality-3-countries).

```{r}
#| label: fig-ncd-mortality-3-countries
#| fig-cap: "Comparing Fiji's NDC mortality rate to France and USA"
#| fig-height: 6
#| fig-width: 8

ncd_mortality |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) + 
  expand_limits(y = 0) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "NCS mortality rates per 100,000 people: Fiji, France, USA",
    x = NULL,
    # y = "Age",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

## Diabetes

```{r}
#| label: prepare-diabetes-data
#| eval: false

# Age-standardized death rates, diabetes mellitus, per 100,000

diabetes_long <- diabetes |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

diabetes_wide <- diabetes_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

# TODO: why only 2004 data?

```

<br>

The following uses 2004 data, which was the latest year available in the GHO data (at least, it's the latest I could find).

At the regional level, Africa is the only region in which the bulk of the countries are suffering a higher rate than the western Pacific (@fig-diabetes-region panel A), while the Americas have the widest range and the worst rates world-wide (panel B).

```{r}
#| label: fig-diabetes-region
#| fig-cap: "Diabetes 2004 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,    ~y,
  "Africa",                  65,   0.027,
  "Americas",                90,   0.007,
  "Eastern Mediterranean",   30,   0.023,
  "Europe",                  15,   0.065,
  "South-East Asia",         36,   0.05,
  "Western Pacific",         58,   0.015,
)

dta_for_plot <- diabetes_long |>
  filter(sex_value == "SEX_BTSX")

ecdf_fun <- ecdf(
  dta_for_plot |>
  filter(parent_location == "Western Pacific") |>
  pull(numeric_value)
  )

fiji_value <- dta_for_plot |>
    filter(spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)

ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- diabetes_long |>
  ggplot() +
   geom_density(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4, 
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "People per 100K",
    y = NULL
  )

p2 <- diabetes_long |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 5, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "People per 100K",
    y = "Percent of all countries in region"
    
  )

p1 + p2 +
  plot_annotation(
    title = glue("Age-standardized death rates, diabetes mellitus, per 100,000 people (2004)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

# TODO: Why isn't the USA point actually on the Americas line?

```

<br>

Fiji ranked 112 out of 191 countries (59th percentile) in death from diabetes (@fig-diabetes). The three-letter country codes used below are detailed in @sec-iso-country-codes.

```{r}
#| label: fig-diabetes
#| fig-cap: "Diabetes 2004"
#| fig-height: 16
#| fig-width: 10

rank_min <- min(diabetes_wide$rank_val)
rank_max <- max(diabetes_wide$rank_val)

diabetes_wide |>
  ggplot() +
  geom_errorbarh(aes(y = spatial_dim_label, xmin = SEX_MLE, , xmax = SEX_FMLE, color = parent_location),
            linewidth = 0.1, height = 0, alpha = 0.8,
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_BTSX, color = parent_location),
            size = 1, alpha = 0.8,
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_MLE, color = parent_location),
            size = 3, alpha = 0.8, shape = "M",
            show.legend = FALSE) +
  geom_point(aes(y = spatial_dim_label, x = SEX_FMLE, color = parent_location),
            size = 3, alpha = 0.8, shape = "F",
            show.legend = FALSE) +
  facet_wrap(~parent_location, scales = "free_y") +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    title = glue("Age-standardized death rates, diabetes mellitus, per 100,000 people (2004)"),
    subtitle = glue("With country rank lowest (1) to highest ({rank_max}) death rate",
                    "\nShowing M, F, and both sexes (the later with a point)"),
    x = NULL,
    y = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

## Prevalence of obesity among adults

The Body Mass Index^[<https://www.cdc.gov/bmi/about/index.html> ], despite it deficiencies, is a useful indicator for comparing population health. A BMI of 30 and above is defined as obese.

```{r}
#| label: prepare-obesity-data

# NCD_BMI_30A Prevalence of obesity among adults, BMI &GreaterEqual; 30 (age-standardized estimate) (%)

obesity_long <- bmi_obesity |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

obesity_wide <- obesity_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

```

<br>

At the regional level the western Pacific experiences the worst obesity rates and the widest range of obesity rates (@fig-obesity-region).

```{r}
#| label: fig-obesity-region
#| fig-cap: "Adult obesity 2022 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- obesity_long |>
  filter(spatial_dim_type == "COUNTRY",
         sex_value == "SEX_BTSX",
         time_dim == 2022)

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,     ~y,
  "Africa",                  15,    0.05,
  "Americas",                37,    0.042,
  "Eastern Mediterranean",   44,    0.028,
  "Europe",                  20.5,  0.072,
  "South-East Asia",         12,    0.078,
  "Western Pacific",         53,    0.014,
)

ecdf_fun <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Western Pacific") |>
    pull(numeric_value)
)

fiji_value <- dta_for_plot |>
    filter(parent_location == "Western Pacific",
           spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(parent_location == "Americas",
           spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)

ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- dta_for_plot |>
  ggplot() +
  geom_density(aes(x = numeric_value, color = parent_location),
               linewidth = 0.6, alpha = 0.8,
               show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 1, size = 4,
            show.legend = FALSE) +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "Percent of adult population",
    y = NULL
  )

p2 <- dta_for_plot |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 3, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 3, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 3, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "Percent of adult population",
    y = "Percent of all countries in region"
    
  )

p1 + p2 +
  plot_annotation(
    title = glue("Adult obesity (2019)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

```

<br>

```{r}
#| label: fig-obesity-fiji-vs-wb-income-group
#| fig-cap: "Adult obesity: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- bmi_obesity |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 1990],
         low_norm = low / low[time_dim == 1990],
         high_norm = high / high[time_dim == 1990],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim),
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "A: Rate",
    x = NULL,
    y = "Rate",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "B: Year 1990 = 100%",
    x = NULL,
    y = "Relative change",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Adult obesity in Fiji has grown more slowly than any World Bank income groups (panel B)",
                 "\nalbeit from a higher starting rate (panel A)"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

```{r}
#| label: fig-obesity-3-countries
#| fig-cap: "Comparing Fiji's adult obesity to France and USA"
#| fig-height: 6
#| fig-width: 8

bmi_obesity |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0.002, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) + 
  expand_limits(y = 0) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Adult obesity: Fiji, France, USA",
    x = NULL,
    y = "Adult obseity",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>

## Mean BMI

The age-adjusted mean BMI provides a similar picture.

```{r}
#| label: prepare-mean-bmi-data

# NCD_BMI_MEAN Mean BMI (kg/m²) (age-standardized estimate)

bmi_mean_long <- bmi_mean |>
  pivot_longer(cols = dim1,
               #names_to = "sex",
               values_to = "sex_value")

bmi_mean_wide <- bmi_mean_long |>
  pivot_wider(#cols = dim1,
               names_from = sex_value,
               values_from = numeric_value) |>
  mutate(rank_val = rank(SEX_BTSX, ties.method = "max"),
         spatial_dim_label = glue("{spatial_dim}: {rank_val}"),
         spatial_dim_label = fct_reorder(spatial_dim_label, rank_val)) |>
  mutate(pct_SEX_MLE = SEX_MLE / 1e5,
         pct_SEX_FMLE = SEX_FMLE / 1e5,
         pct_SEX_BTSX = SEX_BTSX / 1e5,)

```

<br>

At the regional level (@fig-bmi-mean-region) the distribution of mean BMI is consistent with the distribution of adult obesity (@fig-obesity-region).

```{r}
#| label: fig-bmi-mean-region
#| fig-cap: "Mean BMI 2016 by region"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- bmi_mean_long |>
  filter(spatial_dim_type == "COUNTRY",
         sex_value == "SEX_BTSX",
         time_dim == 2016)

dta_for_plot_labels <- tribble(
  ~parent_location,          ~x,     ~y,
  "Africa",                  24.25,    0.27,
  "Americas",                28,    0.40,
  "Eastern Mediterranean",   26.75,    0.19,
  "Europe",                  27.5,    0.55,
  "South-East Asia",         21.5,  0.33,
  "Western Pacific",         29,    0.11,
)

ecdf_fun <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Western Pacific") |>
    pull(numeric_value)
)

fiji_value <- dta_for_plot |>
    filter(parent_location == "Western Pacific",
           spatial_dim == "FJI") |>
    pull(numeric_value)

fiji_percentile <- ecdf_fun(fiji_value)

ecdf_fun_americas <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Americas") |>
    pull(numeric_value)
)

usa_value <- dta_for_plot |>
    filter(parent_location == "Americas",
           spatial_dim == "USA") |>
    pull(numeric_value)

usa_percentile <- ecdf_fun_americas(usa_value)

ecdf_fun_europe <- ecdf(
  dta_for_plot |>
    filter(parent_location == "Europe") |>
    pull(numeric_value)
)

fra_value <- dta_for_plot |>
    filter(parent_location == "Europe",
           spatial_dim == "FRA") |>
    pull(numeric_value)

fra_percentile <- ecdf_fun_europe(fra_value)

p1 <- dta_for_plot |>
  ggplot() +
  geom_density(aes(x = numeric_value, color = parent_location),
               linewidth = 0.6, alpha = 0.8,
               na.rm = TRUE,
               show.legend = FALSE) +
  geom_text(data = dta_for_plot_labels,
            aes(x, y, label = parent_location, color = parent_location),
            hjust = 0, nudge_x = 0, size = 4,
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0.002, 0))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "A: Density",
    x = "Mean BMI",
    y = NULL
  )

p2 <- dta_for_plot |>
  ggplot() +
  stat_ecdf(aes(x = numeric_value, color = parent_location),
            linewidth = 0.6, alpha = 0.8,
            pad = FALSE,
               na.rm = TRUE,
            show.legend = FALSE) +
  annotate("point", x = fiji_value, y = fiji_percentile) +
  annotate("label", x = fiji_value + 0.5, y = fiji_percentile, label = "Republic of Fiji",
            hjust = 0, size = 4) +
  annotate("point", x = usa_value, y = usa_percentile) +
  annotate("label", x = usa_value + 0.5, y = usa_percentile, label = "USA",
            hjust = 0, size = 4) +
  annotate("point", x = fra_value, y = fra_percentile) +
  annotate("label", x = fra_value + 0.5, y = fra_percentile, label = "France",
            hjust = 0, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0.002, 0))) +
  expand_limits(y = 0) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(panel.grid.major.x = element_line(linewidth = 0.03)) +
  labs(
    subtitle = "B: Cumulative Distribution",
    x = "Mean BMI",
    y = "Percent of all countries in region"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Mean BMI (2016)"),
    subtitle = glue("Both sexes, country distribution, by region"),
    caption = glue("Daniel Moul. Source: WHO GHO",
                   "\n(not weighted by population)")
  )

```

<br>

```{r}
#| label: fig-bmi-mean-fiji-vs-wb-income-group
#| fig-cap: "Mean MBI: Fiji and World Bank income groups"
#| fig-height: 6
#| fig-width: 10

dta_for_plot <- bmi_mean |>
  filter(spatial_dim_type == "WORLDBANKINCOMEGROUP" | (spatial_dim_type == "COUNTRY" & spatial_dim == "FJI"),
         dim1 == "SEX_BTSX") |>
  mutate(spatial_dim = factor(spatial_dim, levels = c("WB_LI", "WB_LMI", "WB_UMI", "WB_HI", "FJI"))) |>
  mutate(numeric_value_norm = numeric_value / numeric_value[time_dim == 1975],
         low_norm = low / low[time_dim == 1975],
         high_norm = high / high[time_dim == 1975],
         .by = spatial_dim)

p1 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = spatial_dim),
            show.legend = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  expand_limits(y = 0) +
  labs(
    x = NULL,
    y = "Rate",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p2 <- dta_for_plot |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value_norm, color = spatial_dim)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  expand_limits(y = 0) +
  labs(
    subtitle = "Year 1975 = 100%",
    x = NULL,
    y = "Relative improvement",
    color = "Fiji and WB\nincome group",
    fill = "Fiji and WB\nincome group"
  )

p1 + p2 +
  plot_annotation(
    title = glue("Mean BMI in Fiji has grown even faster",
                 " than World Bank lower and lower middle income groups"),
    subtitle = "Both sexes",
    caption = "Daniel Moul. Source: WHO GHO"
  ) +
  plot_layout(guides = "collect")

```

<br>

While France and the USA have reduced the rate of mean BMI growth, the rate is still increasing in Fiji.

```{r}
#| label: fig-bmi-mean-3-countries
#| fig-cap: "Comparing Fiji's mean BMI to France and USA"
#| fig-height: 6
#| fig-width: 8

bmi_mean |>
  filter(spatial_dim %in% c("FJI", "USA", "FRA")) |>
  ggplot() +
  geom_line(aes(x = time_dim, y = numeric_value, color = dim1)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.002, 0.02))) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) + 
  expand_limits(y = 22) +
  facet_wrap(~spatial_dim) +
  labs(
    title = "Age adjusted mean BMI: Fiji, France, USA",
    x = NULL,
    y = "Mean BMI",
    color = NULL,
    fill = NULL,
    caption = "Daniel Moul. Source: WHO GHO"
  )

```

<br>
